Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation

### 2024-04-18 paper/ms.tex
Updates from Overleaf

% Please add a significance statement to explain the relevance of your work
\significancestatement{Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements. Under the Markov assumption on system dynamics, previous work has provided methods of performing inference for these models. In particular, prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics. We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation. Our new method outperforms previous approaches on a challenging problem in epidemiology.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{K.T. and E.L.I. designed research; K.T. analyzed data; K.T., G.H. and E.L.I. performed research, K.T., G.H., and E.L.I. wrote the manuscript.}
\authordeclaration{The authors declare no competing interests.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}

@@

\begin{abstract}
  Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods. Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques due to (1) the requirement for simulation-based inference that does not require access to the system's transition probabilities and (2) the issue that widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters. We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff. Further, we develop optimization algorithms suited to the Monte Carlo properties of the derivative estimate, including a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation. Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from iterated filtering can beat current state-of-the-art methods on a challenging scientific benchmark problem.
\end{abstract}

@@

\firstpage{4}
% Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting. In this example, there are a total of 11 paragraphs on the first page, counting the first level heading as a paragraph. The value {12} represents the number of the paragraph starting the second page. If a paragraph runs over onto the second page, include a bracket with the paragraph line number starting the second page, followed by the paragraph number in curly brackets, e.g. "\firstpage[4]{11}".

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{M}any scientific models involve highly nonlinear stochastic dynamic systems, where there may be significant uncertainty in both the process dynamics and the measurements of the system. Despite the ubiquity of such partially observed stochastic processes with the Markov property (also known as, continuous-state continuous-time hidden Markov models (HMMs), partially-observed Markov processes (POMPs), or partially-observed stochastic nonlinear dynamical systems) in fields as diverse as epidemiology \cite{he10, stocks17}, ecology \cite{knape12}, and finance \cite{kim08, breto14}, performing estimation and inference within this class of models is a difficult problem that faces several challenges.

Many approaches to inference in these scenarios (such as the EM algorithm, the various Kalman filter variants, and MCMC) either struggle with intractable likelihood functions when the models are complex, or assume access to the probability density of next states given the current state. This is a problem in some critical applications, such as disease modeling, where the models are complex enough such that even obtaining the transition density is an intractable problem. However, the particle filter with the bootstrap proposal, a popular method for solving the filtering problem in partially-observed dynamical systems, provides an unbiased estimate of the likelihood \cite{delmoral2004feynman} without requiring evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.

A series of related algorithms \cite{welch2009abc, wood2010sl, doucet2010pmcmc, ionides08, ionides15} enable parameter estimation within these models. In particular, the improved iterated filtering algorithm (IF2) of Ionides et al. \cite{ionides15} and the particle MCMC algorithm of Andrieu et al. \cite{doucet2010pmcmc} are the only simulation-based full-information maximum likelihood and fully Bayesian methods available in the literature thus far. However, accurate parameter estimation in the presence of significant Monte-Carlo noise in the likelihood estimate remains a significant challenge. Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small. For example, though in practice the algorithm of Ionides et al. \cite{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood. Ionides et al. \cite{Ionides_mcap} attempt to address the third problem, albeit with a computationally-expensive approach.

@@

Recent advances in automatic differentiation (AD) for particle filters \cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased} have drawn attention to AD as an alternative tool for inference in partially observed stochastic processes with the particle filter. However, these existing approaches either are not yet compatible with the desire for simulation-based inference, or are computationally expensive. In particular, existing approaches are either asymptotically biased \cite{blei2018vsmc, jon2018diffpf}, have high variance \cite{poyiadjis11, scibior21}, computationally expensive \cite{corenflos21, chen24}, or require transition densities \cite{poyiadjis11, scibior21, doucet2022particlebased, chen24}.

In particular, the work of Scibior and Wood \cite{scibior2021dpf} demonstrates that the estimators derived in Poyiadjis et al. \cite{poyiadjis11} can be attained with AD, albeit with a simple tweak, that they call the stop-gradient trick, to the particle filter that does not modify its likelihood estimates. Though their method is not fully compatible with simulation-based inference and the mathematical motivation behind the stop-gradient trick is not apparent, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.

\subsection{Our Contributions}
We develop a new approach to statistical inference via AD for particle filters which addresses these concerns. We construct a new theoretical framework that addresses the seemingly incompatible paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.

The key observation is that one can turn this into a problem of differentiating through a series of smooth measurement density ratios and (through the reparametrization trick) a differentiable simulator. This is done through the construction of a suitably reweighted family of particle filters that we call \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$), for some $\alpha \in [0,1]$ that optimizes a bias-variance tradeoff. This family of algorithms encompasses the special case of the biased gradient estimator of \cite{blei2018vsmc} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$, and does not require access to the transition densities -- only requiring access to a differentiable simulator.

@@

In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering. Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1, the Liu-West filter \cite{liuwest01}, and particle MCMC \cite{doucet2010pmcmc}) on a challenging scientific benchmark problem, the Dhaka cholera model of \cite{king08}.

@@

Consider an unobserved Markov process $(X_t)_{t \geq t_0}$, with observations $Y_1,...,Y_N$ realized at values $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$. The process is parameterized by an unknown parameter $\theta \in \Theta \subseteq \R^p$, where the states $X_t$ take values in the state space $\gX \subseteq \R^d$, the observations $Y_n$ take values in $\gY,$ and we write $X_n := X_{t_n}$. By a similar decomposition to that in \cite{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
\begin{align}
    &f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)\\
    &=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).
\end{align}

We call $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ the process model, writing $\texttt{process}\left(x_n; \theta\right)$ for the simulator corresponding to the above process model, $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ the measurement model, and again write $y_n^*$ for the actual values of the observations that were observed. As a matter of terminology, we call $f_{X_{1:N}|Y_{1:N}}(x_{1_N}|y_{1:N}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.

The above are all defined on a single probability space $(\Omega, \Sigma, \prob),$ where $\omega \in \Omega$ denotes an element of the sample space that can be thought of as a random seed. For notational simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$ 
In the rest of this paper, we will identify the seed with an element of the sample space. We do so because the seed determines the entire sequence of pseudo-random numbers as generated by a computer, to maintain an algorithmic and practical interpretation of our theory. When two algorithms share the same seed, this corresponds to the method of common random numbers.

@@

As mentioned earlier, instead of directly differentiating through a particle filter, one can instead differentiate through the simulator and a series of measurement density ratios. We construct an algorithm called \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) in Algorithm \ref{alg:mop} that targets the posterior and likelihood, and whose gradient yields the estimators of \cite{poyiadjis11, scibior21} and \cite{blei2018vsmc} as special cases.

\subsection{Algorithm Outline} 
The broad idea is to run the particle filter twice under the same seed. We first run the particle filter under $\phi$, and then obtain a coupled set of particles by running a bootstrap filter with the process model at $\theta$ but the resampling indices from the previous run at $\phi$. We can then reweight the conditional likelihoods by a correction factor accumulated over time of the ratio of the measurement density under $\theta$ and the measurement density under $\phi$ to account for the ("off-policy") resampling under $\phi$. That is, when we write $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ as shorthand for the measurement density and estimate the conditional likelihood under $\phi$ in the usual way with $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$, we can estimate the conditional likelihood under $\theta$ as either
\begin{equation}
     \label{eq:mop-conditional-likelihood}
     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}} \text{ or } L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
\end{equation}
where each particle is recursively weighted according to 
\begin{equation}
    \label{eq:weighting-scheme}
    w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha,  w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}, w^{F,\theta}_{0,j}= 1,
\end{equation}
and $k_j$ are the indices from resampling according to the measurement density at $\phi$, $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$. The before-resampling conditional likelihood estimate $L_n^{B,\theta,\alpha}$ should be used in general, as it has slightly lower variance than the after-resampling estimate $L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$, but the latter is useful in deriving properties of the MOP-$\alpha$ gradient estimate such as that in Theorem \ref{thm:mop-functional-forms}.

@@

\begin{algorithm}[H]
	\caption{Measurement Off-Policy-$\alpha$}
    \label{alg:mop}
	     \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$. \newline
        \textbf{If $\theta \neq \phi$:}
            Run a particle filter once with parameters $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$, with seed $\omega$. \newline
        \textbf{Else:} Set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$ in the rest of the algorithm. \newline
		\textbf{Initialize } filter particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$ \newline
		\textbf{For} $n=1,...,N$: \newline
            \hspace*{4mm} Accumulate discounted weights, $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$.\newline
            \hspace*{4mm} Simulate process model,
            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$. \newline
            \hspace*{4mm} Measurement density,
            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$. \newline
            \hspace*{4mm} Compute $L_n^{B,\theta,\alpha} ={\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$. \newline
            \hspace*{4mm} Conditional likelihood under $\phi$,
            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$.\newline
            \hspace*{4mm} Select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$. \newline
            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$. \newline
            \hspace*{4mm} Resample corrected weights
            $w_{n,j}^{F,\theta}= w^{P,\theta}_{n,k_j}  { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}$.\newline
            \hspace*{4mm} Compute $ L_n^{A,\theta,\alpha} = L_n^\phi\cdot {\sum_{j=1}^J w^{F,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$.\newline
		\textbf{Return:} likelihood estimate $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}_{j=1}^J.$
\end{algorithm}

@@

As mentioned earlier, the correction factors accumulate over time. If $\theta$ and $\phi$ coincide, one can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood at $\theta$. We do this in order to optimize a bias-variance tradeoff for the score estimate yielded by differentiating through the log-likelihood estimate provided by MOP-$\alpha$. When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, recovering the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}. When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{blei2018vsmc}. $\alpha$ therefore controls an exponentially-weighted moving average, determining how much "memory" the resulting gradient estimate has over ancestral trajectories. This is illustrated in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.

@@

\section{Theoretical Analysis of MOP-$\alpha$}

We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood. To do so, we state the definition of targeting and present three lemmas, whose proof can be found in the supplementary material.
\begin{defn}[Targeting]
    \label{defn:targeting}
    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$,
    \begin{equation}
        E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
    \end{equation}  
        A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$, as $J\to\infty$
    \begin{equation}
        \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X)).
    \end{equation}
\end{defn}

>>>>

% Please add a significance statement to explain the relevance of your work
\significancestatement{Many scientific models involve highly nonlinear stochastic dynamical systems which can be observed only via noisy and incomplete measurements. Prior to this work, iterated filtering algorithms were the only class of algorithms for maximum likelihood estimation that did not require access to the system's transition probabilities, instead needing only a simulator of the system dynamics. We leverage recent advances in automatic differentiation to propose a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation. Our new method outperforms previous approaches on a challenging problem in epidemiology.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{K.T. and E.L.I. designed research; K.T. analyzed data; K.T., G.H. and E.L.I. performed research, K.T., G.H., and E.L.I. wrote the manuscript.}
\authordeclaration{The authors declare no competing interests.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: kevtan@wharton.upenn.edu}

@@

\begin{abstract}
  Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods. Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques as widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters. Furthermore, most approaches are not compatible with the desire for simulation-based inference without access to the system's transition probabilities. We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff. Further, we develop optimization algorithms suited to the Monte Carlo properties of the derivative estimate, including a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation. Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from iterated filtering can beat current state-of-the-art methods on a challenging scientific benchmark problem.
\end{abstract}

@@

\firstpage{5}
% Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting. In this example, there are a total of 11 paragraphs on the first page, counting the first level heading as a paragraph. The value {12} represents the number of the paragraph starting the second page. If a paragraph runs over onto the second page, include a bracket with the paragraph line number starting the second page, followed by the paragraph number in curly brackets, e.g. "\firstpage[4]{11}".

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{M}any scientific models involve highly nonlinear stochastic dynamic systems, where there may be significant uncertainty in both the process dynamics and the measurements of the system. Despite the ubiquity of such partially observed stochastic processes with the Markov property (also known as, continuous-state continuous-time hidden Markov models (HMMs), partially-observed Markov processes (POMPs), or partially-observed stochastic nonlinear dynamical systems) in fields as diverse as epidemiology \cite{he10, stocks17}, ecology \cite{knape12}, and finance \cite{kim08, breto14}, estimation and inference within this class of models is a difficult problem with several challenges.

Popular methods for estimation and inference in these scenarios (such as the EM algorithm, the various Kalman filter variants, and MCMC) either struggle with complex models that have intractable likelihood functions, or assume access to state transition densities. This is a problem in some critical applications, such as disease modeling, where transition densities are unavailable. However, in such cases a simulator may be readily available. In this case, the particle filter with the bootstrap proposal, a popular method for solving the filtering problem in partially-observed dynamical systems, provides an unbiased estimate of the likelihood \cite{delmoral2004feynman} without requiring evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.

A series of related algorithms \cite{welch2009abc, wood2010sl, doucet2010pmcmc, ionides08, ionides15} enable parameter estimation within these models. In particular, the improved iterated filtering algorithm (IF2) \cite{ionides15} is the only simulation-based full-information maximum likelihood method available, whereas particle MCMC \cite{doucet2010pmcmc} and $\text{SMC}^2$ \cite{chopin13} are the only simulation-based full-information Bayesian methods available in the literature thus far. However, accurate parameter estimation with significant Monte-Carlo noise in the likelihood estimate is challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small. For example, though in practice the algorithm of Ionides et al. \cite{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood. This necessitates a new approach.

@@

Recent advances in automatic differentiation (AD) for particle filters \cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased} have drawn attention to AD as a tool for inference in partially observed stochastic processes. However, existing approaches are either asymptotically biased \cite{blei2018vsmc, jon2018diffpf}, have high variance \cite{poyiadjis11, scibior21}, computationally expensive \cite{corenflos21, chen24}, or need access to transition densities \cite{poyiadjis11, scibior21, doucet2022particlebased, chen24}.

In particular, Scibior and Wood \cite{scibior21} show that the estimators derived in Poyiadjis et al. \cite{poyiadjis11} can be attained with AD, through with a simple tweak (the "stop-gradient trick") to the particle filter that does not modify its likelihood estimates. However, their method is not fully compatible with simulation-based inference and the mathematical motivation behind the stop-gradient trick is not apparent.

%We show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.

\subsection{Our Contributions}
We develop a new approach to statistical inference via AD for particle filters which addresses the above concerns. We construct a new theoretical framework that bypasses the paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.

Our key observation is that one can turn this into a problem of differentiating through a series of smooth measurement density ratios and (through the reparametrization trick) a differentiable simulator. To do so, we construct a suitably reweighted family of particle filters that we call \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$), for some $\alpha \in [0,1]$ that optimizes a bias-variance tradeoff. This family encompasses the biased gradient estimator of \cite{blei2018vsmc} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$. It does not require access to the transition densities, and works with just a differentiable simulator.

@@

This allows us to develop improved algorithms for maximum likelihood/MAP estimation in highly nonlinear stochastic dynamical systems. We propose a hybrid algorithm called \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts iterative first or second-order gradient methods with a coarse solution obtained from a few rounds of iterated filtering. Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1 and the Liu-West filter \cite{liuwest01}) on a challenging problem in epidemiology, the Dhaka cholera model of \cite{king08}.

These improvements also extend to Bayesian inference, as we show that we can use the MOP-$\alpha$ gradient estimates within a No-U-Turn Sampler (NUTS) \cite{homan14} in conjunction with a nonparametric empirical Bayes-style prior estimated with IF2 to reduce the burn-in period of particle MCMC \cite{doucet2010pmcmc} from the $1.4 \times 10^6$ iterations in \cite{fasiolo16} to just $500$.

@@

Consider an unobserved Markov process $(X_t)_{t \geq t_0}$, with observations $Y_1,...,Y_N$ realized at values $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$. The process is parameterized by an unknown parameter $\theta \in \Theta \subseteq \R^p$, where the states $X_t$ take values in the state space $\gX \subseteq \R^d$, the observations $Y_n$ take values in $\gY,$ and we write $X_n := X_{t_n}$.

% By a similar decomposition to that in \cite{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
% \begin{align*}
%     &f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)\\
%     &=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).
% \end{align*}

We write $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ for the process model, writing $\texttt{process}\left(x_n; \theta\right)$ for its corresponding simulator, $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ the measurement model, and $y_n^*$ for the actual values of the observations that were observed. We call $f_{X_{1:n}|Y_{1:n}}(x_{1:n}|y_{1:n}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$. Superscripts $x_{n,j}^A$ denote the ancestral trajectory of particle $j$ at time $n$, and $a(\cdot)$ is the ancestor function. For simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$

The above are defined on a probability space $(\Omega, \Sigma, \prob).$ 
To maintain an algorithmic and practical interpretation of our theory, we identify the seed with an element of the sample space $\omega \in \Omega$, as a random seed determines the entire sequence of pseudo-random numbers as generated by a computer.

%When two algorithms share the same seed, this corresponds to the method of common random numbers.

@@

As mentioned earlier, instead of directly differentiating through a particle filter, one can instead differentiate through the simulator and measurement densities. We construct an algorithm called \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) that targets the posterior (i.e. almost sure convergence of every measurable bounded functional of particles to its expectation under the posterior) and likelihood, and whose gradient yields a score estimate encompassing those of \cite{poyiadjis11, scibior21, blei2018vsmc}.

\begin{algorithm}[H]
	\caption{MOP-$\alpha$}
    \label{alg:mop}
	     \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$. \newline
        \textbf{If $\theta \neq \phi$:}
            Fix $\omega$, and filter at $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$. \newline
        \textbf{Else:} Fix $\omega$, and set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$. \newline
		\textbf{Initialize } particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, weights $w^{F,\theta}_{0,j}= 1$. \newline
		\textbf{For} $n=1,...,N$: \newline
            \hspace*{4mm} Accumulate discounted weights, $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$.\newline
            \hspace*{4mm} Simulate process model,
            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$. \newline
            \hspace*{4mm} Measurement density,
            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$. \newline
            \hspace*{4mm} Compute $L_n^{B,\theta,\alpha} ={\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$. \newline
            \hspace*{4mm} Conditional likelihood under $\phi$,
            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$.\newline
            \hspace*{4mm} Select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$. \newline
            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$. \newline
            \hspace*{4mm} Resample corrected weights
            $w_{n,j}^{F,\theta}= w^{P,\theta}_{n,k_j}  { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}$.\newline
            \hspace*{4mm} Compute $ L_n^{A,\theta,\alpha} = L_n^\phi\cdot {\sum_{j=1}^J w^{F,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$.\newline
		\textbf{Return:} likelihood estimate $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, filtering distributions $\{(X_{n,j}^{F, \theta}, w^{F,\theta}_{n,j})\}_{n,j=1}^{N,J}.$
\end{algorithm}

\subsection{Algorithm Outline} 
We obtain two coupled sets of particles, one under $\phi \in \Theta$, and another with the process model at $\theta \in \Theta$ but the resampling indices from the run at $\phi$ (hence the "off-policy" moniker). If $\theta$ and $\phi$ coincide, one only needs one particle filter run at $\theta=\phi$, otherwise one needs two runs at the same seed $\omega \in \Omega$.

We can then reweight the conditional likelihoods by a correction factor accumulated over time to account for the resampling under $\phi$. That is, writing $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ for the measurement density, $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$ for the conditional likelihood estimate under $\phi$, and $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$ for the resampling indices under $\phi$, we can estimate the conditional likelihood under $\theta$ as
\vspace*{-2.5mm}
\begin{equation}
     \label{eq:mop-conditional-likelihood}
     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}} \text{ or } L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
     \vspace*{-2.5mm}
\end{equation}
recursively weighting each particle by 
\vspace*{-2.5mm}
\begin{equation}
    \label{eq:weighting-scheme}
    w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha,  w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}, w^{F,\theta}_{0,j}= 1.
    \vspace*{-2.5mm}
\end{equation}

The before-resampling conditional likelihood estimate $L_n^{B,\theta,\alpha}$ should be used in general, as it has slightly lower variance than the after-resampling estimate $L_n^{A,\theta,\alpha}$, but the latter is useful in deriving properties of the MOP-$\alpha$ gradient estimate such as that in Theorem \ref{thm:mop-functional-forms}.

@@

@@

As mentioned earlier, the correction factors accumulate over time. If $\theta$ and $\phi$ coincide, one can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood at $\theta$. This lets us optimize a bias-variance tradeoff for the MOP-$\alpha$ score estimate. When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, recovering the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}. When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{blei2018vsmc}. $\alpha$ therefore controls an exponentially-weighted moving average, determining how much "memory" the resulting gradient estimate has over ancestral trajectories. This is illustrated in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.

@@

\subsection{Summary of Theoretical Guarantees}

We bypass the issue of differentiating through a Monte Carlo algorithm with discontinuous resampling by turning it into a problem of differentiating through a simulator and a series of measurement density ratios. Although we defer the theoretical analysis of the resulting algorithm, MOP-$\alpha$, to Section \ref{sec:thms}, we highlight a few key results below. MOP-$\alpha$ targets the posterior and likelihood (Theorem \ref{thm:mop-targeting}), yields the estimators of \cite{poyiadjis11, scibior21, blei2018vsmc} as special cases when differentiated (Theorem \ref{thm:mop-functional-forms}) where MOP-$1$ in particular is consistent for the score (Theorem \ref{thm:mop-grad-consistency}), has rates for its bias and variance under different choices of $\alpha$ (Theorem \ref{thm:mop-biasvar}), and enjoys a linear rate of convergence for gradient descent with the resulting gradient estimate (Theorem \ref{thm:mop-convergence}).
